---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
### December 4th, 2016 
### Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

Dataset: Activity monitoring data [52K] The variables included in this dataset are:

steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)

date: The date on which the measurement was taken in YYYY-MM-DD format

interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## Loading and preprocessing the data
```{r}
setwd("D:/nirmala/Coursera-Data Science/Reproducible Research/data")
getwd()
activity <- read.csv("activity.csv",
colClasses = c("numeric", "character","integer"))
summary(activity)
```

## What is mean total number of steps taken per day?
```{r}
library(ggplot2)
total.steps <- tapply(activity$steps, activity$date, na.rm = TRUE, FUN=sum)
total.steps
qplot(total.steps, xlab="Total Number of Steps/day", binwidth=1000)
```
```{r}
mean(total.steps, na.rm = TRUE)
median(total.steps, na.rm = TRUE)
```
## What is the average daily activity pattern?
```{r}
library(ggplot2)
averages <- aggregate(x = list(steps = activity$steps), by = list(interval = activity$interval),
                      FUN = mean, na.rm = TRUE)
ggplot(data = averages, aes(x = interval, y = steps)) +
  geom_line() + 
  xlab("5-minute interval") +
  ylab("average number of steps taken")

```
## Imputing missing values
Missing Values introduce bias in computations - Let's see how many missing values are there
``` {r}
incomplete <- is.na(activity$steps)
table(incomplete) 
```

Let us fill the missing values with a mean value of 5 minute interval
``` {r}
fill.value <- function(steps, interval) {
    filled <- NA
    if (!is.na(steps))
        filled <- c(steps)
    else
        filled <- (averages[averages$interval==interval, "steps"])
    return(filled)
}
filled.activity <- activity
filled.activity$steps <- mapply(fill.value, filled.activity$steps, filled.activity$interval)
```

Now let's create the same histogram with the imputed data

``` {r} 
total.steps <- tapply(filled.activity$steps, filled.activity$date, FUN = sum)
qplot(total.steps, binwidth = 1000, xlab = "total number of steps taken each day")
```

And calcuate the mean and median

```{r}
mean(total.steps)
median(total.steps)
```
## Are there differences in activity patterns between weekdays and weekends?

Using the completed data, let's identify the day of the week for the measurements. 

```{r}
weekday.or.weekend <- function(date) {
    day <- weekdays(date)
    if (day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
        return("weekday")
    else if (day %in% c("Saturday", "Sunday"))
        return("weekend")
    else
        stop("invalid date")
}
filled.activity$date <- as.Date(filled.activity$date)
filled.activity$day <- sapply(filled.activity$date, FUN=weekday.or.weekend)
```
Now, let’s make a plot containing plots of average number of steps taken on weekdays and weekends.

```{r}
averages <- aggregate(steps ~ interval + day, data=filled.activity, mean)
ggplot(averages, aes(interval, steps)) + geom_line() + facet_grid(day ~ .) +
    xlab("5-minute interval") + ylab("Number of steps")
```
